---
title: "DP_HW2"
author: "Maha Siva"
date: "2026-09-21"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(readr)

x <- read_csv("image1.csv")

## Convert data frame into a matrix
x <- data.matrix(x)
```

```{r, echo=FALSE, include=FALSE}
dim(x)
x[1:5, 1:5]

```
## Image 1
```{r}
pca_rank_k_approx <- function(X, k) {
  pca <- prcomp(X)

  Z <- pca$x
  U <- pca$rotation

  Zk <- Z[, 1:k, drop = FALSE]
  Uk <- U[, 1:k, drop = FALSE]

  Xk_centered <- Zk %*% t(Uk)

  Xk <- sweep(Xk_centered, 2, pca$center, FUN = "+")

  error <- norm(X - Xk, type = "F")

  list(
    approx = Xk,
    error = error,
    U = U,
    Z = Z
  )

}


```
### Part 2

```{r}


# Original image
image(
  x,
  col = gray.colors(256),
  main = "Original",
  axes = FALSE
)

# PCA approximations k = 1 through 10
for (k in 1:10) {
  
  result <- pca_rank_k_approx(x, k)
  
  image(
    result$approx,
    col = gray.colors(256),
    main = paste("k =", k),
    axes = FALSE
  )
}


```


```{r}
### Approximation Error 
p <- ncol(x)

errors <- numeric(p)

for (k in 1:p) {
  result <- pca_rank_k_approx(x, k)
  errors[k] <- result$error
}

plot(
  1:p,
  errors,
  type = "l",
  xlab = "k",
  ylab = "Frobenius Error",
  main = "Approximation Error vs. k"
)

```

```{r}
## First Eigenvector and First PC score 
pca <- prcomp(x)

u1 <- pca$rotation[, 1]
z1 <- pca$x[, 1]

plot(
  u1,
  type = "h",
  main = "First Eigenvector",
  xlab = "Variable / Column",
  ylab = "Loading Value"
)

plot(
  z1,
  type = "l",
  main = "First PC Score",
  xlab = "Observation / Row",
  ylab = "PC Score"
)

```


## Interpretation for Image 1

Image 1: The first eigenvector contains both positive and negative loadings across the image columns, with some columns contributing more strongly than others. This indicates that the first principal component captures variation across several parts of the image rather than being dominated by only one column. The first PC scores fluctuate around zero across the rows, showing that the strength and direction of this pattern varies throughout the image.

For Image 1, I think k=10 provides a good approximation. The rank-10 image viusally preserves the main structure of the original, even though the frobenius error continues to decrease at larger values of k.


## image 2
```{r}
# Original image


x2 <- read_csv("image2.csv")

## Convert data frame into a matrix
x2 <- data.matrix(x2)



image(
  x2,
  col = gray.colors(256),
  main = "Original Image 2"
)

# First 10 PCA approximations
for (k in 1:10) {
  result <- pca_rank_k_approx(x2, k)

  image(
    result$approx,
    col = gray.colors(256),
    main = paste("Rank", k, "Approximation")
  )
}


```

## im2 approx error
```{r}
p2 <- ncol(x2)
errors2 <- numeric(p2)

for (k in 1:p2) {
  result <- pca_rank_k_approx(x2, k)
  errors2[k] <- result$error
}

plot(
  1:p2,
  errors2,
  type = "l",
  xlab = "k",
  ylab = "Frobenius Error",
  main = "Image 2: Approximation Error vs. k"
)



```

## Im2, first eigenvector and PC score 
```{r}
pca2 <- prcomp(x2)

u1_2 <- pca2$rotation[, 1]
z1_2 <- pca2$x[, 1]

plot(
  u1_2,
  type = "h",
  main = "Image 2: First Eigenvector",
  xlab = "Variable / Column",
  ylab = "Loading Value"
)

plot(
  z1_2,
  type = "l",
  main = "Image 2: First PC Score",
  xlab = "Observation / Row",
  ylab = "PC Score"
)

```
## analysis:

The first eigenvector shows a clear downward pattern across the columns, indicating that the contribution of the columns to the first principal component changes systematically across the image. The first PC scores fluctuate around zero, showing that the strength and direction of this dominant pattern varies across the rows of the image. 

I think k=2 provides a good approximation for Image 2. The matrix has rank 2, so two principal components are enough to reconstruct the image almost exactly. The Frobenius error after k=2 is essentially zero.


# image 3 
```{r}

x3 <- read_csv("image3.csv")
x3 <- data.matrix(x3)

image(
  x3,
  col = gray.colors(256),
  main = "Original Image 3"
)

for (k in 1:10) {
  result <- pca_rank_k_approx(x3, k)

  image(
    result$approx,
    col = gray.colors(256),
    main = paste("Rank", k, "Approximation")
  )
}

```

## image 3 - error plot 

```{r}


p3 <- ncol(x3)
errors3 <- numeric(p3)

for (k in 1:p3) {
  result <- pca_rank_k_approx(x3, k)
  errors3[k] <- result$error
}

plot(
  1:p3,
  errors3,
  type = "l",
  xlab = "k",
  ylab = "Frobenius Error",
  main = "Image 3: Approximation Error vs. k"
)

```

## first eigenvector and PC score 

```{r}

pca3 <- prcomp(x3)

u1_3 <- pca3$rotation[, 1]
z1_3 <- pca3$x[, 1]

plot(
  u1_3,
  type = "h",
  main = "Image 3: First Eigenvector",
  xlab = "Variable / Column",
  ylab = "Loading Value"
)

plot(
  z1_3,
  type = "l",
  main = "Image 3: First PC Score",
  xlab = "Observation / Row",
  ylab = "PC Score"
)


```


## Analysis

The first eigenvector shows that the contribution of the image columns is not uniform. Some regions have larger negative loadings than others. The first PC score also changes smoothly across the rows. So the dominant image pattern varies systematically across the different vertical regions of the image. 

k=10 might be a good approximation for image 3. The first few ranks are really blurry but by rank 10, the major features of the image are recognizable. The error plot also shows the largest reduction in reconstruction error occurs at small values of k.


## Image 4 

```{r}
x4 <- read_csv("image4.csv")
x4 <- data.matrix(x4)

image(
  x4,
  col = gray.colors(256),
  main = "Original Image 4"
)

for (k in 1:10) {
  result <- pca_rank_k_approx(x4, k)

  image(
    result$approx,
    col = gray.colors(256),
    main = paste("Rank", k, "Approximation")
  )
}



```

### Error plot 

```{r}
p4 <- ncol(x4)
errors4 <- numeric(p4)

for (k in 1:p4) {
  result <- pca_rank_k_approx(x4, k)
  errors4[k] <- result$error
}

plot(
  1:p4,
  errors4,
  type = "l",
  xlab = "k",
  ylab = "Frobenius Error",
  main = "Image 4: Approximation Error vs. k"
)
```
### first eigenvector and PC score 

```{r}
pca4 <- prcomp(x4)

u1_4 <- pca4$rotation[, 1]
z1_4 <- pca4$x[, 1]

plot(
  u1_4,
  type = "h",
  main = "Image 4: First Eigenvector",
  xlab = "Variable / Column",
  ylab = "Loading Value"
)

plot(
  z1_4,
  type = "l",
  main = "Image 4: First PC Score",
  xlab = "Observation / Row",
  ylab = "PC Score"
)


```
## Analysis

The first eigenvector varies across the columns, with the largest negative loadings occurring approximately between columns 60 and 90. This suggests that this region contributes particularly strongly to the dominant variation captured by the first principal component. The first PC scores also vary systematically across the rows, with a particularly large negative region around rows 140–155. This indicates that the dominant pattern represented by PC1 is especially pronounced in that portion of the image.

I think k=10 also provides a good approximation for Image 4 because, at this rank, much of the major structure of the original image has been recovered. The error plot shows a rapid decrease at small values of k, followed by progressively smaller improvements as more components are added.

